---
title: "hw4"
author: "Ty Darnell"
date: "October 2, 2018"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(magrittr)
library(knitr)
pku=read_table("HW4_PKU.txt",col_names=c("pair","case","sibling"),skip=1)
knitr::opts_chunk$set(echo = TRUE)
```

#Problem 1

##Part a

```{r}
differences=pku$case-pku$sibling
```

`case`= the iq of the group of children with PKU
 
`control`= the iq of the group of normal siblings of closest age 

`diff`= `case`-`control`

$\mu_{diff}=$ population mean of the differences

A suitable null and alternative hypotheses is:

$H_0:\mu_{diff} =0$

$H_A: \mu_{diff} \neq 0$

##Part b

```{r}
sm=mean(differences)
se=sd(differences)/sqrt(21)
T= (sm-0)/se
crit.t=qt(.975,20)
pvalue=2*pt(T,20)
```

The sample mean of the differences is $\bar{y}$= `r sm`

The standard error of the differences is se = `r se`

$\alpha= .05$

n=21

n-1 degrees of freedom=20

Test statistic= $t=\dfrac{\bar{y}-\mu}{se}$=`r T`

Critical t=`r crit.t`

pvalue=`r pvalue`

Critical Region: {t: |t| > `r crit.t`}

|`r T`|>`r crit.t`

Since pvalue < .05 and test statistic is in the critical region

Reject $H_0$

##Part c

```{r}
me= crit.t * se
ci=c(sm-me,sm+me)
```

margin of error=me=`r me`

cofidence interval =ci=$(\bar{y}-me,\bar{y}+me)$=`r ci`

##Part d

We reject the null hypothesis since the p-value is less than the significance level and the test statistic lies inside of the critical region and $\mu_0$ is not in the confidence interval. This means that the mean of the differences between the `case` and `control` groups is statistically significant.

The 95% confidence interval for the population mean is `r ci`

This means we are 95% confident that the popluation mean lies within this interval

##Part e

We are assuming the IQ data is normally distributed, all of the observations are independent of each other. Since we have normal data, unknown variance and not n large, we used a t-test.


##Part f

```{r}
index=0:21
sorted_differences=sort(differences)
differences2=c(index,sorted_differences)
smed=sorted_differences[11]
(prob=kable(tibble(r=index,prob=pbinom(index,21,.5)),align="l"))
pval=2*pbinom(6,21,.5)
```

Sample median = $\hat{\zeta_{.5diff}}=$`r smed`

$\zeta_{.5diff}$=population median of the differences of `case` and `control` IQ.

$H_0:\zeta_{.5diff}=\zeta_{.5,0}$

$H_A: \zeta_{.5diff}\neq \zeta_{.5,0}$

R=number of observations $> \zeta_{.5,0}$

r=5 

n-r=16

Critical Region = $C_{\alpha}=\{r:\{0,1,2,3,4,5\}\leq r_{.025} \ \{16,17,18,19,20,21\}\geq r_{.975}\}$

r=11 (the sample median) is not in $C_{.05}$

p-value = `r pval`

`r pval` > .05 and $\hat{\zeta_{.5diff}}$ is not in the critical region so do not reject $H_0$

## Part g

In d) we rejected $H_0$ and in f) we did not reject $H_0$.

The sign test is non parametric and makes fewer assumptions where as the t-test assumes normality and is parametric. For a small sample size it can be argued that the sign test is a better choice since the assumption of normality is a strong assumption especially based on such a small sample size. It is very possible that in this case, the t-test is not a good approximation.

# Problem 2

A Wilcoxon Rank Sum test is appropriate in this case, because we have two small samples sizes with unknown population variance and no assumption of normality.

```{r}
Hyper=c(1100,1320,1350,1450,1600,1850,1900,1990,2050,2120,2200,2210,2500,2610,2720)
Normal=c(1000,1220,1300,1400,1555,1600,1780,1780,1900,2020,2350,2375)
dat <- data.frame(group=c(rep("Hyper",each=15),rep("Normal",each=12)),avg.na=c(Hyper,Normal))

 (dat.rank=dat%>% arrange(avg.na)%>%mutate(rank=(rank(avg.na,ties.method = "average"))))
```
Hyper = the average daily sodium intake of each subject in the hypertension group.

Normal = the average daily sodium intake of each subject in the normal group.

Null hypothesis, the distributions of the two samples are the same.

Alternative hypothesis, the distributions of the two samples are different.

$H_0: \Delta =0$

$H_A: \Delta \neq 0$

```{r}
hyper.ranksum=dat.rank%>%filter(group=="Hyper")%>%summarise(ranksum=sum(rank))
normal.ranksum=dat.rank%>%filter(group=="Normal")%>%summarise(ranksum=sum(rank))
varw=(12*15*28)/12-(12*15*18)/(12*27*26)
Ew=12*28/2

z=(normal.ranksum-Ew)/sqrt(varw)

pval2=2*pnorm(-1.3668861,0,1)
```
$W_1$=`r normal.ranksum`

$Var(W_1)=\dfrac{n_1*n_2(N+1)}{12}-
\dfrac{n_{1}n_{2}}{12N(N-1)}\sum_{i=1}^{q}t_i(t_i-1)(t_i+1)$

$Var(W_1)=$ `r varw`

$E(W_1)=\dfrac{n_1(N+1)}{2}$

$E(W_1)=$ `r Ew`

$Z=\dfrac{W_1 -E(W_1)}{\sqrt{Var(W_1)}}$

Z=`r z `

p-value=`r pval2`



$C_{.05}$ two-tailed $=\{z:|z|>1.96\}$

`r z` is not in the critical region.

since `r pval2` >.05  and our z score is not in the critical region, do not reject $H_0$

Thus we can't conclude that the distributions of the Hypertension and Normal samples are different.

Checking the pvalue using the built in Wilcox test in r we get the same result.

```{r}
wilcox.test(Hyper,Normal,alternative = "two.sided",exact=F,correct=F)

```
