---
title: "hw5"
author: "Ty Darnell"
date: "October 10, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = F,message=F)
library(tidyverse)
library(knitr)
library(epitools)
pgedat=read_table2("HW2_PGE.txt",col_names=c("patient", "ipge", "ca", "hypercalcemia"))
pgedat$hypercalcemia=as.factor(pgedat$hypercalcemia)
```

#Problem 1

$F_1$ = group of subjects without hypercalcemia

$F_2$ = group of subjects with hypercalcemia

$F_1(y)$ = distribution of `ipge` for group of subjects without hypercalcemia

$F_2(y)$ = distribution of `ipge` for group of subjects with hypercalcemia

$H_0:F_1(y)=F_2(y)$ for all y

$H_A: F_1(y)\neq F_2(y)$ for at least one y

$D=\max_y|F_{1n}(y)-F_{2m}(y)|$

$F_{1n}(y)$= Emperical Distribution Function for the sample from $F_1$

$F_{2m}(y)$= Emperical Distribution Function for the sample from $F_2$

```{r}
hca2=pgedat%>%filter(hypercalcemia==1)

hca1= pgedat%>%filter(hypercalcemia==0)

f1=hca1$ipge

f2=hca2$ipge

 ks.test(f1,f2)

```

Using $\alpha = .05$

D=.62727

pvalue =.03242

since $pvalue<.05$ reject $H_0$

We conclude that the distributions of `ipge` may be different for people with and without hypercalcemia.

#Problem 2

```{r}
fisher.test(matrix(c(21,2,2,8),nrow=2),alternative="two.sided")

```
$\pi_1$= Proportion of people dead within 5 years who recieve 1-4 courses of chemotherapy

$\pi_2$= Proportion of people dead within 5 years who recieve 10 or more courses of chemotherapy.

$H_0: \pi_1=\pi_2$

$H_A: \pi_1 \neq \pi_2$

$\alpha=.05$

$pvalue=.0001255$

Reject $H_0$ since $pvalue<.05$

Conclude there is a statistically significant assiociation in the table.

#Problem 3

##Part a


```{r}
church <- matrix(c(89,30514,38,240207),ncol=2,byrow=TRUE)
colnames(church) <- c("D","Dbar")
rownames(church) <- c("E","Ebar")
church <- kable(church)
church
```

Where $E = church < 1$ per week

$\bar{E}=church \geq 1$ per week

$D=death$

$\bar{D}=no \ death$

$P(Death)=\dfrac{127}{54848}=.0023$

$P(Church Attendance) =\dfrac{24245}{54848}=.4420$

$P(church and death)=\dfrac{38}{54848}=.0007$

$P(no church and death)=\dfrac{89}{54848}=.0016$

$p_1=P(death|<1 church)=\dfrac{P(<1 church and death)}{P(<1 church)}=\dfrac{.0016}{.5580}=.0029$

$p_2=P(death|church)=\dfrac{P(church and death)}{P(church)}=\dfrac{.0007}{.4420}=.0016$

Relative risk of an arteriosclerotic death in the 3 year follow up period if one usually attends church less than one week compared to once a week or more.

```{r}
rr=round(.0029/.0016,digits=4)
```


$RR = \dfrac{p_1}{p_2}=\dfrac{.0029}{.0016}=$ `r rr`

This means that a person who attends church less than once a week is 1.8125 times more likely to have an arteriosclerotic death within the 3 year follow up period than someone who attends church at least once a week.

##Part b

```{r}
or=round((.0029/(1-.0029))/(.0016/(1-.0016)),digits = 4)
a=1.96*sqrt(1/89+1/38+1/30514+1/24207)
ci=c(or*exp(-a),or*exp(a))
```


$\hat{OR}=\dfrac{p_1/(1-p_1)}{p_2/(1-p_2)}=\dfrac{.0029/(1-.0029)}{.0016/(1-.0016)}$ = `r or`

95% CI for $\hat{OR}$

$z_{\alpha/2}=1.96$

$\hat{OR}*e^{\pm 1.96\sqrt{1/89+1/38+1/30514+1/24207}}=$ (`r ci`)

##Part c

```{r}
pe=100*(or-rr)/rr
```

Percent error of $\hat{OR}=100(OR-RR)/RR=$ `r pe`%

#Problem 4

##Part a

```{r}
X<-array(c(7,55,31,269,7,20,18,112,7,33,24,114,40,88,45,172,
            34,50,24,55,27,55,24,58,30,34,17,17),
          dim=c(2,2,7),
          dimnames=list(CupsCoffee=c("GE5","LT5"),MIcase=c("Yes","No"),Smoking=c("Never","Former","1-14","15-24","25-34","35-44","45+")))
rows <- dim(X)[1]
cols <- dim(X)[2]
result <- matrix(0, nrow=rows, ncol=cols)
for (i in seq(rows)) {
  for (j in seq(cols)) {
    result[i,j] <- sum(X[i,j,])
  }
}
rownames(result)=c(">=5","<5")
colnames(result)=c("MI","Control")
kable(result)
```

Using R to collapse table 6.11 over the smoking categories yields the same results as the table in problem 6.13.


##Part b

```{r}
p1=round((152/1467)/(335/1467),digits=4)
p2=round((335/1467)/(1132/1467),digits=4)
or2=(p1/(1-p1))/(p2/(1-p2))
b=1.96*sqrt(1/152+1/183+1/335+1/797)
ci2=c(or2*exp(-b),or2*exp(b))
```

$E=\geq 5$ cups of coffee per day

$\bar{E}= <5$ cups of coffee per day

$D=MI$

$\bar{D}=Control$

__Odds ratio and 95% CI ignoring smoker status__

$p_1=P(MI| \geq 5 cups)=\dfrac{P(\geq 5 cups and MI)}{P(\geq 5 cups)}=\dfrac{152/1467}{335/1467}$= `r p1`

$p_2=P(MI|<5 cups)=\dfrac{P(<5 cups and MI)}{P(<5 cups)}=\dfrac{335/1467}{1132/1467}$= `r p2`

$\hat{OR}=\dfrac{p_1/(1-p_1)}{p_2/(1-p_2)}=$ `r or2`

95% CI for $\hat{OR}$

$z_{\alpha/2}=1.96$

$\hat{OR}*e^{\pm 1.96\sqrt{1/152+1/183+1/335+1/797}}=$ (`r ci2`)

```{r}
nosmoke=oddsratio.wald(result)$measure
nosmoke[c(0,2),]
```


__Odds ratio and 95% CI taking into account smoker status__

```{r}
never=oddsratio.wald(X[,,1])$measure
former=oddsratio.wald(X[,,2])$measure
s1.14=oddsratio.wald(X[,,3])$measure
s15.24=oddsratio.wald(X[,,4])$measure
s25.34=oddsratio.wald(X[,,5])$measure
s35.44=oddsratio.wald(X[,,6])$measure
s45=oddsratio.wald(X[,,7])$measure
```

```{r}
never[c(0,2),]
former[c(0,2),]
s1.14[c(0,2),]
s15.24[c(0,2),]
s25.34[c(0,2),]
s35.44[c(0,2),]
s45[c(0,2),]
```

##Part c

```{r}
mantelhaen.test(X)
```
$H_0: OR =1$

$H_A: OR \neq 1$

Since pvalue < .05 reject $H_0$

There is evidence that smoking status may confound the association between coffee drinking and MI.
