---
title: "hw8"
author: "Ty Darnell"
date: "November 26, 2018"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(knitr)
library(data.table)
library(magrittr)
library(survival)
library(ggfortify)
library(reshape2)
knitr::opts_chunk$set(echo = TRUE,warning=F, message=F)
#options(scipen=999)
```

# Problem 1

## Part a

```{r}
cnames=c("status", "N.total", "N.h", "N.c", "n.total", "n.h", "n.c")
dat=fread("HW8_Q1.txt")
colnames(dat)=cnames
```

Crude Death Rate per 100,000 operations

$\dfrac{100,000*deaths}{operations}$

 halothane

$\dfrac{2375}{146200}*100,000=1624.487$ deaths per 100,000 operations

 cyclopropane
 
 $\dfrac{2109}{68169}*100,000=3093.782$ deaths per 100,000 operations
 
 Using a Z-test for proportions to determine if the two rates are significantly different.
 
 $\alpha=.05$
 
 $H_0: \pi_1= \pi_2$
 
 $H_A: \pi_1 \neq \pi_2$
 
```{r}
prop.test(x=c(2375,2109), n=c(146200,68169), alternative = "two.sided")
```
 
Since p-value < $\alpha$ reject $H_0$ 

Conclude there is a significant differnce between the two rates.

## Part b

```{r}
dat%<>%mutate(p.h=n.h/N.h,p.c=n.c/N.c)

dat2=dat %>%
mutate(varProph = p.h*(1-p.h)/N.h,
varPropc = p.c*(1-p.c)/N.c,
weights = N.total/sum(N.total))

ans=dat2%>% summarize(stdProph = sum(p.h*weights),
          stdPropc = sum(p.c*weights),
          adjDiff = stdProph-stdPropc,       
          varAdjDiff = sum(weights^2*(varProph+varPropc))/sum(weights)^2, 
          Z = (adjDiff)/sqrt(varAdjDiff),          
          pValue = pnorm(Z)) 

stdh=ans$stdProph*100000
stdc=ans$stdPropc*100000
```

Standardized Halothane = `r stdh`

Standardized Cyclopropane = `r stdc`

Using a Z-test to determine if the two rates are significantly different.
 
 $\alpha=.05$
 
 $H_0: \pi_{1_{adj}}= \pi_{2_{adj}}$
 
$H_A: \pi_{1_{adj}} \neq \pi_{2_{adj}}$
 
```{r}
kable(ans)
```
 
 Reject $H_0$ since p-value$<\alpha$
 
 Conclude the two standardized rates are not equal.
 
## Part c
 
```{r}
dati=dat %>% 
mutate(refPop = n.total/N.total,
exp=N.h*refPop)
ans2=dati%>% summarize(o = sum(n.h), 
            e=sum(exp),          
            s = o/e,          
            varO = o,         
            varE = sum((N.h/N.total)^2*n.total),
            varS = (varO+s^2*varE)/e^2,   
            Z = (s-1)/sqrt(varS),
            pVal = pnorm(Z))
```
 
```{r}
kable(ans2)
o=ans2$o
e=ans2$e
```
 Using a Z-test to test whether $\pi_{halothane}/\pi_{overall}$=1
 
 $\alpha=.05$
 
 $\pi_{halothane}=$ `r o`
 
 $\pi_{overall}=$ `r e`
 
 Standardized Incidence Ratio =$s=\dfrac{O}{E}$
 
 =$\pi_{halothane}/\pi_{overall}$=`r ans2$s`

$H_0:\pi_{halothane}/\pi_{overall}=1$

$H_A:\pi_{halothane}/\pi_{overall}\neq 1$

$Var(s)=\dfrac{Var(O)+s^2Var(E)}{E^2}$ = `r ans2$varS`

Since `r ans2$pVal` $<\alpha$

Reject $H_0$ conclude $\pi_{halothane}/\pi_{overall}\neq 1$

# Problem 2

## Part a

```{r}
sur=fread("HW8_SURV.txt")
cnames2=c("id","timedeath","death","age","group")
colnames(sur)=cnames2
```

```{r}
sur2=sur%>%arrange(timedeath)
fit <- survfit(Surv(timedeath, death)~group,data=sur2)
autoplot(fit,xlab="t",ylab="S(t)",main = "Kaplan-Meier Curves for Placebo and Treatment Groups")
summary(fit)
```

## Part b

Log rank test to test whether the distribution of survival times
is the same in the two groups.

$\alpha=.05$

$H_0: S_1(t)=S_2(t) \ \forall t$

$H_A: S_1(t)\neq S_2(t)$ for at least one t

```{r}
survdiff(Surv(timedeath, death)~group,data=sur2)
```

p-value=.003<$\alpha$

Reject $H_0$ since p-value< $\alpha$, thus $S_1(t)\neq S_2(t)$ for at least one t

## Part c

Using the proportional hazards model to test whether the distribution of survival times is the same in the two groups.

$\beta$=-0.6972

$\alpha=.05$

$H_0: \beta =0$

$H_A: \beta \neq 0$


```{r}
 phaz=coxph(Surv(timedeath, death)~group,data=sur2)
phaz
```

pvalue=.00358<$\alpha$

Reject $H_0$ conclude the $\beta$ coefficient is significantly different from 0. Thus the distribution of survival times is not the same in the two groups.

## Part d

```{r}
conf=exp(confint(phaz))
```


Estimate for hazard ratio for group 2 relative to group 1= 0.4980

95% CI for the true hazards ratio: `r conf`

## Part e

```{r}
 phaz1=coxph(Surv(timedeath, death)~group+age,data=sur2)
phaz1
```

Adding age to the proportional hazards model

p-value for age=0.03387

since p-value < $\alpha$, age does have a signifcant effect on survival.

$\beta$ for group adding in age= 0.48639

Adjusting for age does not make a substantial difference to the estimate of the treatment effect since the group coefiecient does not change substantially.

## Part f

Median survival time for placebo group=197

S(t)=.5 at t=197

```{r}
sur2%>%filter(group==1)%>%summarize(median=median(timedeath))
```

