---
title: "2017ty"
author: "Ty Darnell"
date: "7/18/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=F,warning=F)
```


```{r}
library(tidyverse)
library(vcd)
library(nortest)
library(epitools)
library(multcomp)
library(data.table)
library(knitr)
```


```{r}
mot=fread("motor.dat")
sub=fread("subtype.dat")
```

replace impossible GA_ultra value with NA

```{r}
mot=mot%>%mutate(GA_ultra=replace(GA_ultra,GA_ultra>70,NA))
mot$PD_post=na_if(mot$PD_post,".") #change missing to NA
```


# Problem 1

## a)

```{r}
round(mean(mot$ppnum),digits=2)
```

mean=1.5

## b)

$\lambda=1.5$ 

$X\sim Pois(1.5)$

$n=453$

$P(X\geq 4)=1-P(X<4)=1-P(X\leq 3)=.0656=\pi$

expected number of women having at least 4 kids= .0656*453=29.74

```{r}
pi=1-ppois(3,1.5)
n=453
u=pi*n
z=0-u/sqrt(u*(1-pi))
z
```

$\alpha=.05$

$H_0: \pi=.0656$

$H_A: \pi\neq .0656$

Under $H_0: Z=\dfrac{Y-n\pi}{n\pi(1-\pi)}\sim N(0,1)$ Where Y is the observed number (0)

z=-5.64 critical z= 1.96

$|-5.64|>1.96$ Thus reject $H_0$ The observed number 0 is much less than the expected number, 29.74

## c)

We will use lilliefors KS test to test whether the normal distribution is a good model for the distribution of GA_ultra

```{r}
lillie.test(mot$GA_ultra)
shapiro.test(mot$GA_ultra)
```

Let Y=GA_ultra

Testing $H_0: Y\sim N(\mu, \sigma^2)$ vs $H_A$: Y is not normally distributed

$pvalue<.05$ thus reject $H_0$, there is evidence that GA_ultra is not normally distributed

## d)

```{r}
ga=mot%>%dplyr::select(GA_ultra,GA_est)
ga$GA_est=cut(ga$GA_est,c(0,37,40,Inf),right=F,labels=c("low","med","high"))
ga$GA_ultra=cut(ga$GA_ultra,c(0,37,40,Inf),right=F,labels=c("low","med","high"))
```


```{r}
gatab=table(ga)
gatab2=addmargins(gatab)
gatab2
```

Measure of agreement kappa

```{r}
(k=Kappa(gatab))
confint(k)
```

kappa=.6898$\approx .69$ 95% CI= (.627,.753)

kappa=.69 indicates moderate agreement between the two versions of GA

## e)

Dichotomize ga_ultra by  <37 weeks to define preterm delivery

preterm = 0 not preterm delivery

preterm = 1 preterm delivery

```{r}
mot2=mot%>%mutate(preterm=as.numeric(GA_ultra<37)) #dichotomize GA_ultra as preterm/ not preterm

```

Conducting a chi square test for trend

Let $p_i$ denote the probability preterm=1 in ppnum category i. i= 0,1,2,3+ 

$H_0: p_0 = p_1 = p_2 = p_3$

$H_A: p_0 \leq p_1\leq p_2 \leq p_3$ or $p_0 \geq p_1\geq p_2 \geq p_3$ with at least one of the inequalities being strict

```{r}
preterm=mot2$preterm
ppnum=mot2$ppnum
tab1=table(preterm,ppnum) #Create 2x2 Disease exposure table
tab2=addmargins(tab1) #adding margins to table
tab2
```

```{r}
chisq.test(tab1)
```

prop.trend.test(x=number of events,n=number of trials,score=group number)
In our case we have a disease exposure model
x= number of diease+
n = column total
score = exposure level

```{r}
x=c(12,12,12,12)
n=c(90,153,102,107)
group=c(1,2,3,4)
prop.trend.test(x,n,group)
```

Do not reject $H_0$ the risk of preterm delivery does not vary monotonically with ppnum. In fact it is very consistent throughout each group.

## f)


## g) 

```{r}

```


