---
title: "662 tests"
author: "Ty Darnell"
date: "7/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=F,warning=F)
```


```{r}
library(tidyverse)
```


# z-stat functions

```{r}
#one sample z-stat function
zstat=function(y,u0=0){
y=na.omit(y)
yb=mean(y)
n=length(y)
s=sd(y)
zstat=(yb-u0)/(s/sqrt(n))
zstat
}

#two sample z-stat function
#NA values must be removed prior to using this
zstat2=function(y1,y2,u0=0){
yb1=mean(y1)
n1=length(y1)
s1=sd(y1)
yb2=mean(y2)
n2=length(y2)
s2=sd(y2)
zstat=((yb1-yb2)-u0)/sqrt(s1^2/n1+s2^2/n2)
zstat
}

#p-value function for alpha=.05
pval=function(z,twosided=T){
  if (twosided==T) {
    return(2*pnorm(z))
  }
  else
    pnorm(z)
}

```



# Mcnemar test stat

```{r}
#takes a 2x2 table (in epid form) and computes the mcnemar test stat
mstat=function(table){
 t2=table[2]
 t3=table[3]
 m=(t2-t3)^2/(t3+t2)
 m
}
```

# Modified Levene Test

Homogeneity of variance – Modified Levene test (Brown-Forsythe test): apply
ANOVA to the absolute deviations from group medians
$d_{ij}=|Y_{ij}-\tilde{Y_i}|$
Use usual F test; rejection indicates lack of homogeneity
(Ordinary Levene test uses means, not medians)
Modified Levene test is robust to normality
does not require equal sample sizes

```{r}
Levene <- function(y, group)
{
group <- as.factor(group) # precautionary
medians <- tapply(y, group, median)
resp <- abs(y - medians[group])
anova(lm(resp ~ group))
}
```


# Epi functions

```{r}
etable=function(exposure,disease){
  t=table(exposure,disease)[2:1,2:1]
  rownames(t)=c('E+','E-')
  colnames(t)=c('D+','D-')
  t
}


riskratio=function(a,b,c,d){
  e=a/(a+b)
  f=c/(c+d)
  c(e,f,e/f)
}


ciriskr=function(ecases,enoncases,uecases,uenoncases){
  a=ecases
  b=enoncases
  c=uecases
  d=uenoncases
  e=a/(a+b)
  f=c/(c+d)
  rr=e/f
  z=c(-1.96,1.96)
  se=sqrt(1/a-1/(a+b)+
            1/c-1/(c+d))
  ci=exp(log(rr)+z*se)
  c("Risk Ratio"=rr,"95 CI"=ci)
}


rateratio=function(cases,noncases,pyrcases,pyrnoncases){
  a=cases
  b=noncases
  p=pyrcases
  q=pyrnoncases
  (a/p)/(b/q)
}


cirater=function(cases,noncases,pyrcases,pyrnoncases){
  a=cases
  b=noncases
  p=pyrcases
  q=pyrnoncases
  rr=(a/p)/(b/q)
  se=sqrt(1/a+1/b)
  z=c(-1.96,1.96)
  ci=exp(log(rr)+z*se)
  c("Rate Ratio"=rr,"95 CI"=ci)
}


cioddsr=function(ecases,enoncases,uecases,uenoncases){
  a=ecases
  b=enoncases
  c=uecases
  d=uenoncases
  or=(a/c)*(d/b)
  se=sqrt(1/a+1/b+1/c+1/d)
  z=c(-1.96,1.96)
  ci=exp(log(or)+z*se)
  c("Odds Ratio"=or,"95 CI"=ci)
}


etable2=function(exposure,disease){
  t=table(exposure,disease)[2:1,2:1]
  rownames(t)=c('E+','E-')
  colnames(t)=c('D+','D-')
  addmargins(t)
}
```