---
title: "662hw6"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = F, warning = F)
```

```{r}
library(data.table)
library(tidyverse)
library(knitr)
library(mosaic)
library(ggfortify)
library(nortest)
library(vcd)
```

```{r}
column_names=c("id","group","trg")
column_classes=c("character","factor","integer")
trgdat=fread("data/trg.txt",colClasses = column_classes)
setnames(trgdat,column_names)
```


```{r}
trgdat%>%group_by(group)%>%summarise(N=n(),mean=round(mean(trg),digits=1),median=median(trg),sd=round(sd(trg),digits=2))
```

Analysis plan:
Anova model (cell means): $Y_{ij}=\mu_i+\epsilon_{ij}, i=1,2,3,4,5$
i=1  0 mg/kg (control group)
i=2  75 mg/kg
i=3  150 mg/kg
i=4  300 mg/kg
i=5  600 mg/kg

$\mu_i$ = population mean trg level in ith doseage group

index j denotes the jth rat in the ith doseage group

$\epsilon_{ij}\overset{iid}{\sim} N(0,\sigma^2)$

The primary interest is in pairwise comparisons between groups, to determine which groups differ from one another. 
The group sizes are unequal, varying from 7 to 10, so it is more
appropriate to use the Scheffe or Bonferroni method to adjust for multiple comparisons.
We will use Scheffe’s method.

```{r}
trg$group=as.factor(trg$group)
fit=aov(trgdat$trg~trgdat$group)
anova(fit)
```

```{r}
plot(fit$fitted.values,fit$residuals)
qq<-qqnorm(fit$residuals)
qqline(fit$residuals)
```
```{r}
cor.test(qq$x,qq$y)
```

```{r}
lillie.test(fit$residuals)
```
pvalue=.121>.05 Thus do not reject $H_0$, the normality assumption of the residuals may have been violated.

```{r}
Levene <- function(y, group)
{
group <- as.factor(group) # precautionary
medians <- tapply(y, group, median)
resp <- abs(y - medians[group])
anova(lm(resp ~ group))
}
Levene(trgdat$trg,trgdat$group)
```

Levene Test, do not reject $H_0$ 

```{r}
scheffe=pairw.anova(trgdat$trg,trgdat$group,conf.level=.95,method="scheffe")
scheffe$summary
```
```{r}
m1=aov(trgdat$trg~trgdat$group)
Anova(m1)
```

