---
title: "hw6"
author: "Ty Darnell"
date: "November 8, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = F, message = F)
library(tidyverse)
library(knitr)
library(data.table)
library(MASS)
library(asbio)
library(nortest)
column_names=c("id","group","trg")
column_classes=c("character","factor","integer")
trgdat=fread("HW6_TRG.txt",colClasses = column_classes)
setnames(trgdat,column_names)

```

#Problem 1

Using a pairwise scheffe Anova Test

Assumptions:

Homogeneity of variance

Normality and independence of residual error

Linearity

$\mu_1$ is the mean triglyceride level of the control group (group 1)

$\mu_2$ is the mean of group 2

$\dots$

$\mu_5$ is the mean of group 5

We have ${5\choose 2}$=10 null and alternative hypotheses represented by i and j each taking values of the 5 groups.

Each $H_0$ is testing if the mean trg levels are equal between two groups

Each $H_A$ is that the mean trg levels are not equal between those two groups

$H_0:\mu_i=\mu_j \quad (i\neq j)$

$H_A: \mu_i \neq \mu_j \quad (i\neq j)$


```{r}
pairwise=pairw.anova(trgdat$trg,trgdat$group,conf.level=.95,method="scheffe")
(critfval=qf(.95,4,41))
```

Looking at the table of result from the Scheffe test we have:

```{r}
kable(pairwise[4])
```

$\alpha=.05$

Conclusion:

Fail to reject $H_0$ for $(\mu_1,\mu_2)$ $(\mu_1,\mu_3)$  $(\mu_2,\mu_3)$ $(\mu_3,\mu_4)$  $(\mu_4,\mu_5)$ based on the p-values provided in the table which are all $>\alpha$

This means that there is not a statistically signifcant difference between the means of trg level for each of these pairwise comparisons

Reject $H_0$ for $(\mu_1,\mu_4)$ $(\mu_2,\mu_4)$  $(\mu_1,\mu_5)$ $(\mu_2,\mu_5)$  $(\mu_3,\mu_5)$
based on the p-values provided in the table which are all $<\alpha$

This means that there is a difference between the means of the trg levels of these pairwise comparisons.

Conducting a global anova F-Test we have:

```{r}
av <- aov(trgdat$trg~trgdat$group)
kable(anova(av))
```

$H_0: \mu_1=\mu_2=\dots=\mu_5$

$H_A:$ at least one inequality

Critical Value for F-statistic with degrees of freedom 4,41 =2.6

$C_{.05}=\{F>F_{4,41,.95}=2.6\}$

MSR=Mean square regression

MSE = Mean square error

Global F test statistic: $F^*$  = MSR/MSE= 15.02

$F^*>$ critical F value

$15.02>2.6$

Conclusion: Reject $H_0$ since p-value< .05 and test statistic is in the critical region.

Conclude that at least one group has a different triglyceride mean.

```{r}
qq1 <- qqnorm(av$residuals); qqline(av$residuals)
cor.test(qq1$x,qq1$y)
lillie.test(av$residuals)
shapiro.test(av$residuals)
```

The residuals appear approximately normally distributed.

conducting a boxcox transformation in order to determine whether a transformation of the data would be appropriate.

```{r}
bc=boxcox(trgdat$trg~trgdat$group)
boxcox(trgdat$trg~trgdat$group)
lambda <- bc$x[which.max(bc$y)]
```

$\lambda$=`r lambda` 

```{r}
fit <- aov(log(trgdat$trg)~trgdat$group)
qq <- qqnorm(fit$residuals,xlab="Fitted Values",ylab="Residuals")
qqnorm(fit$residuals,xlab="Fitted Values",ylab="Residuals"); qqline(fit$residuals)
cor.test(qq$x,qq$y)
lillie.test(fit$residuals)
shapiro.test(fit$residuals)
```

Looking at the qqplots of the original and transformed data, there does not appear to be an improvement in normality with the transformed data. Thus a transformation is not appropriate.

# Problem 2

##Part i)

Running an anova F test to determine if there is a difference between the trg mean of the control group and the trg mean of the combined treatment groups.

Assumptions:

Homogeneity of variance

Normality and independence of residual error

Linearity

$\mu_0$ = mean of control groups

$\mu_1$ = mean of combined treatment groups

$H_0: \mu_0=\mu_1$

$H_A: \mu_0 \neq \mu_1$

```{r}
trgdat2=trgdat%>%mutate(group2=factor(as.integer(group!="1")))
av2 <- aov(trgdat2$trg~trgdat2$group2)
kable(anova(av2))

qf(.95,1,44)

```

$\alpha=.05$

p-value = .0003

Critical Value for F-statistic with degrees of freedom 1,44 =4.0617

F test statistic: $F^*$  = MSR/MSE= 15.4457

$F^*>$ critical F value

$15.4457>4.0617$

Conclusion: Reject $H_0$ since p-value< .05 and test statistic is in the critical region.

Conclude that the control group and treatments groups have different a triglyceride mean.

##Part ii)

Running a Pearson correlation test to determine whether there is a linear relationship between group number and trg level.

We will look at the correlation between group and trg level.

Assumptions:

Homogeneity of variance

Normality and independence of residual error

Linearity

$H_0:$ Correlation = 0

$H_A:$ Correlation $\neq$ 0


```{r}
(cortest=cor.test(as.numeric(trgdat$group),trgdat$trg))
```

p-value = `r cortest$p.value`< .05

Conclusion: Reject $H_0$ since the p-value < $\alpha$. Conclude the true correlation is not equal to 0 thus there is a linear relationship between group and trg level.

A correlation of -.768 represents a negative linear relationship between group and trg level. 

#Problem 3

```{r}
trgdat$dose=recode(trgdat$group,"1"=0,"2"=75,"3"=150,"4"=300,"5"=600)
trgdat3=trgdat %>% filter(dose!=0)
(model=lm(trgdat3$trg~trgdat3$dose))
control=trgdat%>%filter(dose==0)
meancontrol=mean(control$trg)
```

Assumptions:

Linearity: $Y_i = \alpha+\beta X_i +\epsilon_i$

Xs are fixed constants

$\epsilon_i$ are independent and identially distributed $\sim N(0,\sigma^2)$

$\hat{Y_i}$=Predicted triglyceride level

$X_i$=dose (mg of peppermint extract)

$\hat{Y_i}=\hat{\alpha} +\hat{\beta} X_i$

$\hat{Y_i}=238.55423+-.05058X$

```{r}
ggplot(data = trgdat3,aes(x=dose,y=trg))+geom_point()+geom_smooth(method="lm")
summary(model)
```

se=standard error=.00676

$\beta$=-.05058

95% confidence interval for slope:

$\beta \pm$ .00676

(-.064<$\beta$<-.037)


Predicted mean triglyceride level for the control group:

$\hat{Y_0}=238.55423-.05058(0)$

$\hat{Y_0}=238.55423$

The observed trg mean of the control group was
`r meancontrol`

Percent Error= $\dfrac{Observed-Predicted}{Predicted}*100$

$\dfrac{244.2-238.55423}{238.55423}*100=2.37\%$

Since the percent error is $2.37\%$ the predicted and sample mean values for the control group are reasonably similiar