---
title: "Power"
output:
  prettydoc::html_pretty:
     theme: cayman
     df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=F,warning=F)
```

```{r,echo=F}
library(data.table)
library(tidyverse)
library(modelr)
library(knitr)
library(matlib)
library(ggfortify)
library(linmod)
```


```{r}
library(pwr)
```


# Method 1: Using exisiting power functions

## two sample t-test

```{r}
effect= pwr.t.test(n = 20, sig.level = 0.05,  power = 0.8, type = "two.sample")
power= pwr.t.test(n = 20, d=0.9, sig.level = 0.05,  type = "two.sample")
samplesize = pwr.t.test(d=0.9, sig.level = 0.05, power= 0.8,  type = "two.sample")
```


```{r,echo=F}
result=c(effect$d,power$power,samplesize$n)
name=c("d","power","n")
kable(rbind(name,round(result,digits=3)),row.names = F)
```


## two sample freq comparison


# Method 2: write own power function


n is the total sample size and design is balanced
mu1 and mu2 are means of the two comparison groups
sigma2 is the variance of the random error
omega is the non-central parameter


```{r}
omega<- function(n, mu1, mu2, sigma2){
  return(n*(mu1-mu2)^2/(4*sigma2))
 }
```


## varing sample size


```{r}
n<- seq(50,1000, by=50) 
n.omega<- omega(n, 0, 0.25, 1)

n.power<- rep(1, length(n.omega)) 
for(k in 1:length(n.omega)){
  f.crit<- qf(0.95, df1=1, df2=n[k]-2, 0)
  n.power[k]<-1-pf(f.crit, df1=1, df2=n[k]-2, n.omega[k])
 }

plot(n, n.power, ylim=c(0,1),xlab="Sample Size", ylab="Power")
points(n, n.power, type="l", lty=1)

```



## varying mu2


```{r}
n.omega<- omega(n, 0, 0.15, 1)

n.power<- rep(1, length(n.omega)) 
for(k in 1:length(n.omega)){
  f.crit<- qf(0.95, df1=1, df2=n[k]-2, 0)
   n.power[k]<-1-pf(f.crit, df1=1, df2=n[k]-2, n.omega[k])
 }

plot(n, n.power, ylim=c(0,1),xlab="Sample Size", ylab="Power")
points(n, n.power, type="l", lty=2)

```



## varying sigma2

```{r}
n.omega<- omega(n, 0, 0.25, 2)

n.power<- rep(1, length(n.omega)) 
for(k in 1:length(n.omega)){
  f.crit<- qf(0.95, df1=1, df2=n[k]-2, 0)
   n.power[k]<-1-pf(f.crit, df1=1, df2=n[k]-2, n.omega[k])
 }

plot(n, n.power, ylim=c(0,1),xlab="Sample Size", ylab="Power")
points(n, n.power, type="l", lty=3)

legend(700, 0.3, lty=c(1,2,3), legend=c("mu2=0.25", "mu2=0.15", "sigma2=2"))
```




# Method 2: run power analysis with simulation




```{r}
mu1=0
mu2=0.25
n=500
sigma2=1
n.sim<- 1000  #number of simulations
n.test<- rep(0, n.sim)

for(k in 1:n.sim){
  x<- c(rep(0,n/2), rep(1,n/2))
  y<- mu1 + mu2*x + rnorm(n, 0, sigma2)
  flm<- lm(y~x)
  n.test[k]<- (summary(flm)$coefficients[2,3])^2
  }
```


```{r}
power = mean(n.test>1.96^2)
```


```{r,echo=F}
kable(power,col.names="Power")

```



